---
title: "Data Wrangling"
author: "Jana Nassir"
date: "11/20/2021"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("tidyverse")
library(readr)
```

```{r}
income <- read_csv("~/finalp/datasets/income.csv")
crime <- read_csv("~/finalp/datasets/crime.csv")
edu_attainment <- read_csv("~/finalp/datasets/edu_attainment.csv")
education_finance <- read_csv("~/finalp/datasets/education_finance.csv")
unemployment <- read_csv("~/finalp/datasets/unemployment.csv")
```

```{r}
income <- filter(income, !is.na(State))
```
Cleaning the data, getting rid of unwanted rows.

```{r}
crime <- rename(crime, state = state_name)
crime$state[is.na(crime$state)] <- "United States"
crime$state_abbr[is.na(crime$state_abbr)] <- "USA"
crime <- subset (crime, select = -c(rape_revised, caveats))
```

To be able to join it with other data later on.

```{r}
crimes <- filter(crime, year>=1984)
```

To align the years with other data sets. Removing columns with mostly NA.

```{r}
yearly_income <- gather(income, year, income, 2:38) 
yearly_income <- rename(yearly_income, state = State)
yearly_income$year <- as.numeric(yearly_income$year)
```

```{r cars}
yearly_unemployment <- gather(unemployment, year, unemployment, 2:40)
yearly_unemployment <- filter(yearly_unemployment, year>=1984)
yearly_unemployment$year <- as.numeric(yearly_unemployment$year)
```

Making the yearly income and unemployment from wide to long such that the data becomes attachable to the main "crimes" table we created. 

```{r}
crime_dataset <- full_join (crimes, yearly_income, by = c("state", "year"))
crime_dataset <- full_join (crime_dataset, yearly_unemployment, by = c("state", "year"))
crime_dataset <- filter(crime_dataset, year >= 1984, year<=2019)
```

This data set includes all the crime data from 1984 to 2019 including the income and unemployment rates.

```{r}
edu_attainment[edu_attainment == "United States (with PR)"] <- "United States"
edu_attainment<-subset(edu_attainment, state!="Puerto Rico")
```

Renaming data to make it compatible with other sets. Removed Puerto Rico since it doesn't exist in any other data sets.

```{r}
crime_dataset_edu <- crime_dataset
crime_dataset_edu <- full_join (crime_dataset_edu, edu_attainment, by = c("state", "year"))
crime_dataset_edu <- filter(crime_dataset_edu, year >= 2008, year <2020)
```

This data set is the same as crime data set but includes the educational attainment of people aged 25-64. The rate represents the number of adults in this age range which have studied after high school. It was made into a seperate table because the data about the education is only from 2008.

```{r}
edu_finance <- education_finance
edu_finance <- rename(edu_finance, year = YEAR, state = STATE)
edu_finance <- full_join(edu_finance, crime_dataset_edu, by = c("state", "year"))
edu_finance <- filter(edu_finance, year >= 2008, year < 2017)
edu_finance <-subset(edu_finance, state!="United States")
```

This table is similar to the one before except that it include the amount of money that the government spends on financing for schools. It was made into another data set because it is only up to 2017. 


```{r}
write.table (crime_dataset, file = "/Users/asus/Documents/finalp/datasets/MODIFIED/crime_dataset.csv", row.names = F, sep = ",")

write.table (crime_dataset_edu, file = "/Users/asus/Documents/finalp/datasets/MODIFIED/crime_dataset_edu.csv", row.names = F, sep = ",")

write.table (crimes, file = "/Users/asus/Documents/finalp/datasets/MODIFIED/crimes.csv", row.names = F, sep = ",")

write.table (yearly_income, file = "/Users/asus/Documents/finalp/datasets/MODIFIED/yearly_income.csv", row.names = F, sep = ",")

write.table (yearly_unemployment, file = "/Users/asus/Documents/finalp/datasets/MODIFIED/yearly_unemployment.csv", row.names = F, sep = ",")

write.table (edu_finance, file = "/Users/asus/Documents/finalp/datasets/MODIFIED/edu_finance.csv", row.names = F, sep = ",")
```

To download the tables. 